
\chapter{Extending \ee}

\section{Introduction}

Current PaaS solutions available on market are well known for their low flexibility. Some of them work only on a specific cloud environment, others only with a few development frameworks. \ee\ tries to overcome this issue by providing an extensible architecture. Although the out-of-box version of \ee\ is quite limited, with some programming is possible to extend it to provide support to new \emph{i)} cloud providers, \emph{ii)} package types, \emph{iii)} service types, and \emph{iv)} node selection policies. By ``extending'' we mean no current \ee\ code need to be changed, and that each new extension can be implemented by the means of a well-defined process, which are now described in this section.

\section{Supporting new cloud providers}

In \ee, cloud providers are just a source of virtual machines provisioning. Any technology able to create new virtual machines may be used as ``cloud provider''. 

To implement a new cloud provider, it is necessary to implement the \textsf{CloudProvider} interface (Listing~\ref{lst:cloud_provider}). Current implementations are \textsf{AWSCloudProvider} (that uses EC2 service), \textsf{OpenStackKeystoneCloudProvider}, and \textsf{FixedCloudProvider} (that always points to the same user-defined VMs). An example of new cloud provider implementation could be the \textsf{VirtualBoxCloudProvider}, that would use VirtualBox on the developer machine to create new VMs (this is an example more suited to test environments).

\lstset{
language=Java,
numbers=left
}

{\footnotesize
\begin{lstlisting}[caption=\textsf{CloudProvider} interface., label=lst:cloud_provider]
public interface CloudProvider {

    public String getCloudProviderName();

    public CloudNode createNode(NodeSpec nodeSpec) 
    	throws NodeNotCreatedException;

    public CloudNode getNode(String nodeId) 
    	throws NodeNotFoundException;

    public List<CloudNode> getNodes();

    public void destroyNode(String id) 
    	throws NodeNotDestroyed, NodeNotFoundException;

    public CloudNode createOrUseExistingNode(NodeSpec nodeSpec) 
    	throws NodeNotCreatedException;

    public void setCloudConfiguration(CloudConfiguration cloudConfiguration);

}
\end{lstlisting}
}

Implementations should use the \texttt{cloudConfiguration} object to retrieve 
configuration properties supplied by EE administrators. Such properties usually encompass
user credentials to access the infrastructure provider service, and options
such as VM types or images. The \texttt{cloudConfiguration} object is injected
into the cloud provider instance by the EE.

\emph{Important note}: in the current implementation, \ee\ is tailored to work with nodes running \emph{Ubuntu 12.04}. Therefore, \textsf{CloudProvider} implementors should provide Ubuntu 12.04 nodes. 

The next step is to edit the \texttt{extensible/cloud\_providers.properties} file, located on EnactmentEngine resources folder. You must add a line in the format \verb!NAME=full.qualified.class.name!, where the key is just an alias that you can freely define (since it does not conflict with other existing keys on the same file), and the value is the full qualified name of the \textsf{CloudProvider} implementing class. It is also necessary to recompile the EnactmentEngine project in such way it can access the implementing class. One suggestion is by adding your class in your local maven repository and edit the EnactmentEngine project's pom to make EnactmentEngine dependent on your project holding the new cloud provider

Finally, to use your new cloud provider, it is necessary to configure the \texttt{clouds.properties},
adding a cloud account whose \verb!CLOUD_PROVIDER! property values the \verb!NAME! defined in the \texttt{cloud\_providers.properties} file.


\section{Supporting new package types}

Services may be delivered in different package types, such as JARs, WARs, etc.
Each package type has its own specific deployment procedures, as well its specific process to start the service.
When using different technologies, such as Python, to write new services, you will need to define a new package type,
as well the deployment procedure associated with it. Such procedure is specified in Chef recipe. 

So, the first step is to create a new Chef cookbook similar to the ``jar'' and ``war'' recipes already provided by \ee.
These cookbooks are actually templates that EE will use to create specific cookbooks to each service to be deployed.
You can use the the constants \$PACKAGE\_URL and \$NAME within your cookbook recipe and attributes files. These constants will be injected by \ee to each specific recipe.
You can have an idea about how to use them by looking to the WAR cookbook implementation, in Listing~\ref{lst:recipe_template} and Listing~\ref{lst:attributes_template}.
After writing the new recipe, you must associate this recipe to the new package type by editing the \texttt{extending/cookbooks.properties} file.

\begin{lstlisting}[frame=trbl, label=lst:recipe_template, caption=Recipe template for WAR deployment.]
include_recipe "apt" 
include_recipe "tomcat::choreos"

remote_file "war_file" do
        source "#{node['CHOReOSData']['serviceData']['$NAME']['PackageURL']}"
        path "#{node['tomcat']['webapp_dir']}/$NAME.war"
        mode "0755"
        action :create_if_missing
end

file "#{node['tomcat']['webapp_dir']}/$NAME.war" do
        action :nothing
end
\end{lstlisting}

\begin{lstlisting}[frame=trbl, label=lst:attributes_template, caption=Attributes template for WAR deployment.]
default['CHOReOSData']['serviceData']['$NAME']['PackageURL'] = "$PACKAGE_URL"
\end{lstlisting}

It is up to EE to ``guess'' the service URI too.
A service URI follows the format \texttt{http://IP:PORT/CONTEXT}.
And the CONTEXT formation rule is package type dependent.
Therefore, when extending package type, it is necessary to create a new \textsf{URIContextRetriever}
implementation and to link this implementation to its package type in the \textsf{URIContextRetrieverFactory} class. To make this relationship, it is enough to add
a single line in the factory, by adding a new entry in the \texttt{classMap} variable.
Both \textsf{URIContextRetriever} and \textsf{URIContextRetrieverFactory} classes are in the \textsf{org.ow2.choreos.services.datamodel.uri} package, on the EnactmentEngineAPI project.

\emph{Hint:} if your package type is based on some kind of container to run the services,
such as Tomcat, it may be a good idea to prepare a new image with this container already
installed and running. So, you can configure EE to create VMs with an already running
instance of your chosen container (e.g. JBoss).
This strategy helps in achieving a faster deployment.


\section{Supporting new service types}

Although web services came to tackle the interoperability issue, today we have a couple of technologies implementing the concept of services.
The main standards in this context are SOAP and REST, but other technologies could be used to implement services, such as JMS.

In the \ee\ context, the service type affects only how the \texttt{setInvocationAddress} is invoked.
Therefore, to support a new service type, you have only to write a new \textsf{ContextSender} (Listing~\ref{lst:context_sender}) implementation.

{\footnotesize
\begin{lstlisting}[caption=\textsf{ContextSender} interface., label=lst:context_sender]
public interface ContextSender {

    public void sendContext(String serviceEndpoint, 
                            String partnerRole, 
                            String partnerName,
                      List<String> partnerEndpoints) throws ContextNotSentException;
}
\end{lstlisting}
}

The final step is to edit the \texttt{extensible/context\_sender.properties} file, located on EnactmentEngine resources folder. You must add a line in the format \verb!SERVICE_TYPE=full.qualified.class.name!, where the key is the name of the new service type, and the value is the full qualified name of the \textsf{ContextSender} implementing class. It is also necessary to recompile the EnactmentEngine project in such way it can access the implementing class. Now you can create new service specs using the just-created service type! But be sure services implementation are prepared to receive the \texttt{setInvocationAddress} invocation.

\section{Supporting new node selection policies}

A node selection policy defines the \emph{mapping} of services to cloud nodes.
Since cloud nodes are dynamically created, node selection policies must be flexible, and not rely on hard-coded IPs.
A node selection policy may define nodes to be used based on services non-functional properties.
To create a new node selector, you must create a new \textsf{NodeSelector} (Listing~\ref{lst:node_selector}) implementation.
Pay attention that such implementation must be thread-safe, since multiple threads will invoke concurrently the method \texttt{select}.

{\footnotesize
\begin{lstlisting}[caption=\textsf{NodeSelector} interface., label=lst:node_selector]
public interface NodeSelector extends Selector<CloudNode, DeployableServiceSpec> {

}

public interface Selector<T, R> {

    public List<T> select(R requirements, int objectsQuantity) throws NotSelectedException;

}
\end{lstlisting}
}

After writing the new node selector, you must associate this selector to a label by editing the \texttt{extensible/node\_selector.properties} file, at EnactmentEngine resources folder.
To use the new selector, finally, you must attribute the defined label to the NODE\_SELECTOR property on the \texttt{ee.properties} file.







